Transfer of Learned Opponent Models in Zero Sum Games

### 2020-01-14 Experiment_2/Experiment_2_modelling.Rmd
draft update discussion

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

@@

```{r}
# Read in cleaned data file
dat_exp2 <- read.csv("data_exp2.csv")

# Function to produce best reponse to ai_action in exp2
exp2_best_response <- function(game,ai_action){
  if (game == "rps"){
    if (ai_action == "R") {return("P")}
    else if (ai_action == "P") {return("S")}
    else if (ai_action == "S") {return("R")}

@@

test_data = subset(dat_exp2,human_id == "bZg1T3MNoItL6b7qAAAj")
exp2_naive_bayes(c(0.5,0.1),test_data,c("level0","level1","level2"),"-2loglik",TRUE)
```
## Fitting Bayesian model with transfer  
```{r}
exp2_TR_bayes <- list()
for(id in unique(dat_exp2$human_id)) {
  exp2_TR_bayes[[id]] <- list()
  tdat <- subset(dat_exp2,human_id == id)

@@

MBM2_modelling <- list()
for(id in unique(dat$human_id)) {
  MBM_modelling[[id]] <- list()
  tdat <- subset(dat,human_id == id)
  # MBM_modelling[[id]] <- optim(c(1,0.1),fn=QMBM,gr = NULL, data=tdat,"-2loglik", gamma =0 , lower = c(0,0), upper = c(10,0.99), method="L-BFGS-B")

@@

```{r}
exp2_table_results <- table(exp2_results[, "condition"],c("random","Bayes Tr","Bayes No Tr", "QL_states","S_EWA")[apply(exp2_results[,c("Random_BIC","Bayes_Tr_BIC","Bayes_No_Tr_BIC","QL_states_BIC","EWA_self_BIC")],1,which.min)])

@@

```{r}
exp2_model_comp <- data.frame()
for(id in unique(dat_exp2$human_id)) {
  tdat <- subset(dat_exp2,human_id == id)
  tot_score <- sum(tdat$score)
  tot_time <- sum(tdat$human_rt)
  early_dat <- subset(tdat,between(tdat$round,2,6) & (game =="fwg"))
  #early_dat <- subset(tdat,between(tdat$round,2,6) & (game =="fwg" | game =="numbers") )
  tr_score <- sum(early_dat$score)
  id_results <- subset(exp2_results, ID == id)
  min_BIC <- apply(id_results[,c("Random_BIC","Bayes_Tr_BIC","Bayes_No_Tr_BIC","QL_states_BIC","EWA_self_BIC")],1,min)

  best_model <- c("random","Bayes Tr","Bayes NT", "QL_states","S_EWA")[apply(id_results[,c("Random_BIC","Bayes_Tr_BIC","Bayes_No_Tr_BIC","QL_states_BIC","EWA_self_BIC")],1,which.min)]
  # 
 exp2_model_comp <- rbind(exp2_model_comp ,
                       data.frame(
                         "human_id" = id,
                         "condition" = dat_exp2[dat_exp2$human_id==id,"condition"][1],
                         "Early_game_score" = tr_score,
                         "Total_score" = tot_score,
                         "Best_model" = best_model,
                         "Total_time" = sum(tdat$human_rt),
                         "TR_minus_NT_BIC" = id_results[,"Bayes_Tr_BIC"] - id_results[,"Bayes_No_Tr_BIC"],
                         "Rand_minus_best_BIC" =  id_results[,"Random_BIC"] - min_BIC

>>>>

```{r setup, include=FALSE}
library(knitr)
knitr::opts_chunk$set(echo = TRUE)

@@

```{r}
# Read in cleaned data file
dat_exp2 <- read.csv("data_exp2.csv")
test_data = subset(dat_exp2,human_id == "bZg1T3MNoItL6b7qAAAj")
```

```{r}
# Function to produce best reponse to ai_action in exp2
exp2_best_response <- function(game,ai_action){
  if (game == "rps"){
    if (ai_action == "R") {return("P")}
    else if (ai_action == "P") {return("S")}
    else if (ai_action == "S") {return("R")}

@@

exp2_naive_bayes(c(0.5,0.1),test_data,c("level0","level1","level2"),"-2loglik",TRUE)
```
## Fitting Bayesian model with transfer  
```{r}
exp2_TR_bayes <- list()
for(id in unique(dat_exp2$human_id)) {
  exp2_TR_bayes[[id]] <- list()
  tdat <- subset(dat_exp2,human_id == id)

@@

MBM2_modelling <- list()
for(id in unique(dat_exp2$human_id)) {
  MBM2_modelling[[id]] <- list()
  tdat <- subset(dat_exp2,human_id == id)
  # MBM_modelling[[id]] <- optim(c(1,0.1),fn=QMBM,gr = NULL, data=tdat,"-2loglik", gamma =0 , lower = c(0,0), upper = c(10,0.99), method="L-BFGS-B")

@@

```{r}
exp2_table_results <- table(exp2_results[, "condition"],c("random","Bayes Tr","Bayes No Tr", "QL_states","S_EWA","MBM")[apply(exp2_results[,c("Random_BIC","Bayes_Tr_BIC","Bayes_No_Tr_BIC","QL_states_BIC","EWA_self_BIC","MBM2_BIC")],1,which.min)])

@@

```{r}
exp2_model_comp <- data.frame()
for(id in unique(dat_exp2$human_id)) {
  tdat <- subset(dat_exp2,human_id == id)
  tot_score <- sum(tdat$score)
  tot_time <- sum(tdat$human_rt)
  early_dat <- subset(tdat,between(tdat$round,2,6) & (game =="fwg"))
  #early_dat <- subset(tdat,between(tdat$round,2,6) & (game =="fwg" | game =="numbers") )
  tr_score <- sum(early_dat$score)
  id_results <- subset(exp2_results, ID == id)
  min_BIC <- apply(id_results[,c("Random_BIC","Bayes_Tr_BIC","Bayes_No_Tr_BIC","QL_states_BIC","EWA_self_BIC","MBM2_BIC")],1,min)

  best_model <- c("random","Bayes Tr","Bayes NT", "QL_states","S_EWA", "MBM")[apply(id_results[,c("Random_BIC","Bayes_Tr_BIC","Bayes_No_Tr_BIC","QL_states_BIC","EWA_self_BIC","MBM2_BIC")],1,which.min)]
  # 
 exp2_model_comp <- rbind(exp2_model_comp ,
                       data.frame(
                         "human_id" = id,
                         "condition" = dat_exp2[dat_exp2$human_id==id,"condition"][1],
                         "Early_game_score" = tr_score,
                         "Total_score" = tot_score,
                         "Best_model" = best_model,
                         "Total_time" = sum(tdat$human_rt),
                         "TR_minus_NT_BIC" = id_results[,"Bayes_Tr_BIC"] - id_results[,"Bayes_No_Tr_BIC"],
                         "Rand_minus_best_BIC" =  id_results[,"Random_BIC"] - min_BIC

### 2020-01-14 Report/draft_report_v1.Rmd
draft update discussion

```{r load-packages, include = FALSE}
library("papaja")
require(knitr)
require(citr)
```

@@

@@

Figure 10 shows the marginal mean scores for each game for both level-1 and level-2 facing players. Graphically we can see that the pattern is dissimilar between level-1 and level-2 players, and we suspect transfer to be positively associated with similarity and negatively with degree of sophistication of the agent. To test these hypotheses, we run statistical tests on scores by game and opponent against the null hypothesis of 0 (no transfer). Table XXXXXXX shows the scores on rounds 2-6 for both type of opponents in each game. We also report the t-stats and p-values to test the null hypothesis that scores on these early rounds are equal to 0.  Rejecting the null hypothesis means that prior learning has been used to beat the opponent in these early rounds.  For level-1 facing players, there is evidence of learning transfer to both FWG and NUMBERS. For level-2 facing players, there is evidence for transfer to the similar game FWG (albeit scores are lower than for level-1 player) but not to the dissimilar game of NUMBERS.

@@

As a measure for learning transfer we will compare scores only on rounds 2-6 of each game (excluding the very first round where play is necessarily random). Because the number of rounds here is very limited, there should be very little learning within games, and we should be better measure learning transfer if it exits. In the figure below, we show the average score across participants in rounds 2-6. These scores are also averaged across the levels of condition (meaning they are irrespective of which opponent players faced first). For both the FWG and SHOOT games, score in the early rounds of the first stage are significantly higher than 0 for both opponent types. (Level-1 opponent: FWG: t(270) = 4.99, p < 0.0001; SHOOT:  t(270) = 6.66, p < 0.0001; Level-2 opponent: FWG: t(270) = 4.40, p < 0.0001; SHOOT:  t(270) = 3.21, p =0.004 ). Looking more specifically at early scores by type of opponent faced, we confirm the result from the first experiment that it is easier to transfer learning to the more dissimilar game (Shoot) when facing a level 1 opponent. Indeed, while the early scores of FWG for level-1 and level-2 facing players are not significantly different from each other, the score of the players facing the level-1 opponent is indeed almost 0.2 point per round higher than that of players facing level-2 opponents, and the difference is statistically significant ( t(144) = 2.45 , p = 0.01). These early scores have also been adjusted to account for the fact that the shootout game has higher average scores when playing randomly, and therefore this difference is really due to better learning transfer and not due to chance.

@@

>>>>

```{r load-packages, include = FALSE}
#library("papaja")
require(knitr)
require(citr)
# using some functions dplyr, ggpubr, PairedData and sjPlot. Need to be loaded. 
library(tidyr)
library(dplyr)
library(MASS)
library(ggpubr)
library(PairedData)
library(multcompView)
library(lsmeans)
```

@@

```{r, include = FALSE}
#Experiment 1 raw data
dat <- read.csv("../data20180719.csv")

# transform 'winner' variable in numeric score
dat$score <- recode(dat$winner, human = 1, tie = 0, ai = -1)
# create a new variable 'block' with round 1...25 = block 1 and round 26...50 as block 2
dat$block <- factor(as.numeric(cut(dat$round,2)),labels =c("first half", "second half"))

# create a new variable "game.f" as a factor variable of games
dat$game.f <- factor(dat$game, labels = c("RPS","FWG","NUM"),levels=c("rps","fwg","numbers"))

#Group data by human_id and calculate mean score per block of each game.
dat2 <- dat %>% 
  group_by(human_id,condition,game,block,game.f) %>% 
      summarize(block_score = mean(score))

# Group data by game and ID
dat3 <- group_by(dat2, human_id,game.f,game) %>% summarise(game_score = mean(block_score))

```{r, include = TRUE}
# AFEX PACKAGE FOR RUNNING RM ANOVA AND POST HOC TESTS ############################
citation(package = "afex", lib.loc = NULL, auto = NULL)
library(afex)
a1 <- aov_car( block_score ~ game.f*block*condition + Error(human_id/(game.f*block)), dat2)
summary(a1)
(a1_nice <- nice(a1, es = attr(a1$anova_table, "es"),
  observed = attr(a1$anova_table, "observed"),
  correction = attr(a1$anova_table, "correction"), MSE = NULL,
  intercept = NULL, p_adjust_method = attr(a1$anova_table,
  "p_adjust_method"), sig_symbols = attr(a1$anova_table, "sig_symbols")))

write.table(a1_nice,file = "a1.txt", sep = ",", quote = FALSE, row.names = F)
```

```{r}
#Pair waise comparison and post hoc tests for the interaction: Game * Block using lsmeans

# game score differences statistically significant?
(ls0 <- lsmeans(a1, "game.f"))
(lsm0 <- update(pairs(ls0, reverse = TRUE), by=NULL, adjust = "holm"))

# Main effect of block 
(ls01 <- lsmeans(a1, "block"))
(lsm01 <- update(pairs(ls01, reverse = TRUE), by=NULL, adjust = "holm"))
lsmip(a1, ~block)

# Main effect of condition
(ls02 <- lsmeans(a1, "condition"))
(lsm02 <- update(pairs(ls02, reverse = TRUE), by=NULL, adjust = "holm"))

```{r}
# Learning transfer using first half scores: Pairwise comparison of first halves by game ( first half RPS vs first half WFG scores....)
(ls1 <- lsmeans(a1, "game.f", by="block"))
(lsm1 <- update(pairs(ls1, reverse = TRUE), by=NULL, adjust = "holm"))
lsmip(a1, game.f ~ block)
```

```{r}
(means.int <- lsmeans(a1, specs = c("game.f","block","condition")))

# Transfer level1: compare H1FWG with H1RPS 
con1  <- contrast(means.int, list(H1_G2vsH1_G1forlvl1 = c(-1,1,0,0,0,0,0,0,0,0,0,0), H1_G3vsH1_G1forlvl1=c(-1,0,1,0,0,0,0,0,0,0,0,0)))
summary(con1, adjust = "holm")

# Transfer level2: compare H1FWG with H1RPS 
con2  <- contrast(means.int, list(H1_G2vsH1_G1forlvl2 = c(0,0,0,0,0,0,-1,1,0,0,0,0), H1_G3vsH1_G1forlvl2=c(0,0,0,0,0,0,-1,0,1,0,0,0)))
summary(con2, adjust = "holm")
```

```{r}

#looking at TRIALS 2 to 6 to test robustness of evidence for transfer of learning of opponent strategy #########

dat_26 <- subset(dat,round >1 & round <7, drop =TRUE)
dat2_6 <- dat_26 %>% 
  group_by(human_id,condition,game.f,confidence,difficulty) %>% 
      summarise(early_score = mean(score))

# Check group means and SDs
group_by(dat2_6, game.f) %>%
  summarise(
    count = n(),
    mean = mean(early_score, na.rm = TRUE),
    sd = sd(early_score, na.rm = TRUE)
    )
group_by(dat2_6, game.f,condition) %>%
  summarise(
    count = n(),
    mean = mean(early_score, na.rm = TRUE),
    sd = sd(early_score, na.rm = TRUE)
    )
# plot scores per game 
ggerrorplot(dat2_6, x = "game.f", y = "early_score", group = 1, color="game.f", desc_stat = "mean_ci",palette = c("#00AFBB", "#E7B800", "#FC4E07"), order = c("RPS", "FWG","NUM"), ylab = "Score", xlab = "Games") 
```

@@

Figure XXXXX shows the marginal mean scores for each game for both level-1 and level-2 facing players. Graphically we can see that the pattern is dissimilar between level-1 and level-2 players, and we suspect transfer to be positively associated with similarity and negatively with degree of sophistication of the agent. To test these hypotheses, we run statistical tests on scores by game and opponent against the null hypothesis of 0 (no transfer). Table XXXXXXX shows the scores on rounds 2-6 for both type of opponents in each game. We also report the t-stats and p-values to test the null hypothesis that scores on these early rounds are equal to 0.  Rejecting the null hypothesis means that prior learning has been used to beat the opponent in these early rounds.  For level-1 facing players, there is evidence of learning transfer to both FWG and NUMBERS. For level-2 facing players, there is evidence for transfer to the similar game FWG (albeit scores are lower than for level-1 player) but not to the dissimilar game of NUMBERS.

@@

As a measure for learning transfer we will compare scores only on the first 5 rounds (rounds 2-6 of each game, excluding the very first round where play is necessarily random). Because the number of rounds here is very limited, there should be very little learning within games, and we should be better measure learning transfer if it exits. In the figure below, we show the average score across participants in rounds 2-6. These scores are also averaged across the levels of condition (meaning they are irrespective of which opponent players faced first). For both the FWG and SHOOT games, score in the early rounds of the first stage are significantly higher than 0 for both opponent types. (Level-1 opponent: FWG: t(270) = 4.99, p < 0.0001; SHOOT:  t(270) = 6.66, p < 0.0001; Level-2 opponent: FWG: t(270) = 4.40, p < 0.0001; SHOOT:  t(270) = 3.21, p =0.004 ). Looking more specifically at early scores by type of opponent faced, we confirm the result from the first experiment that it is easier to transfer learning to the more dissimilar game (Shoot) when facing a level 1 opponent. Indeed, while the early scores of FWG for level-1 and level-2 facing players are not significantly different from each other, the score of the players facing the level-1 opponent is indeed almost 0.2 point per round higher than that of players facing level-2 opponents, and the difference is statistically significant ( t(144) = 2.45 , p = 0.01). These early scores have also been adjusted to account for the fact that the shootout game has higher average scores when playing randomly, and therefore this difference is really due to better learning transfer and not due to chance.

# Computational modelling

To gain more insight into how participants played the games against the computer opponents, we compared multiple models of strategies the players may have been using to learn how to beat the opponent.  As a base model, we assume play is random, and each potential action is chosen with equal probability. Note that this corresponds to the Nash equilibrium strategy. Next, we include for comparison purposes a few other basic models of behavior, such as Q-learning on action choices, as well as a more sophisticatd Q-learning model incorporating a state space consisting of last round play. The Bayesian approach is represented by models whereby agents would update an initial belief over the opponent's strategy consisting of various level-k strategies (resp. 0,1 and 2). We fitted two models in which agents best respond to the opponent model. In the first, the beliefs about the opponent's adopted strategy are transfered across games, and another in which no transfer of opponent models is allowed.

Next, we fitted a self-tuning EWA model (Ho et al., 2004) both on action choices without states and with states consisting of lat round play. The self-tuning EWA model generalizes both reinforcement learning as well as fictitious play models and has been proven to do well than equilibrium models in various game situations. Finally, we also included a model-based reinforcement learning approach. whereby the agent learns the transition probabilities between states ans uses that to choose the best action, given the current state and it's belief about what the new state will be.

All models were fit to each participant in both experiments, with optimal parameters being estimated using maximum likelihood. Using information criteria based Bayesian model comparison (BIC), the best fitting model for each participant was chosen and we compared the number of participants whose behavior was explained by each model. Table XXX  shows the results for experiment 1, while Table XXXX shows those for experiment 2. we can see that while some participant's learning behavior was either random or explained by some of the base models, a significant number of participants in experiment one had learning most consistent with Q-learning conditional on last round play. In experiment 2 hwever, that model wasn't as successful as the bayesian approached in fitting player's action choices. In experiment 2 when participants faced both level-1 and level-2 agents sequentially, the bayesian models (with or without transfer) were by far the most successful.

Next we compared the performance of players whose actions are consistent with each of our hypothesized models. Figure XXXX shows the average cumulative performance of players across games, for participants grouped by which model best fits their behavior. participants whose actions are most consistent with learning an opponent model in a Bayesian way had the best overall performance (both with and witout transfer), followed by Q-learning conditional on last round play. EWA, QL and random players had, understandably the lowest performance.

@@

In this study, we investigated human learning transfer across games by making human participants play against computer agents with rule-based level-k strategies. We were interested in uncovering evidence for transfer and exploring whether it is modulated by the degree of similarity between games and the sophistication of the agent.

The results of our first online experiment show that the majority of participants learn to adapt to the opponent strategy over multiple interactions and generalise this learning to the similar game. We found that using results on very early rounds is a better proxy for measuring transfer as it is not tainted by any within game learning. Using this approach, we showed that transfer to the more dissimilar game was modulated by the degree of sophistication of the agent, with evidence for transfer when players face the less sophisticated agent but not the more sophisticated one.

In the second online experiment, the paradigm was better suited to explore our main research question, related to learning transfer rather than within-game learning. Indeed, when facing two opponents sequentially, there are many more opportunities to test transfer than before. Indeed, there are 2 opportunities to transfer knowledge within each game, and a total of three games, which means 6 opportunities to test transfer. When we made players face only one type of opponent each, we only had two possible learning transfers to test. In that regard, the results on learning transfer confirmed prior findings from the first experiment. While there was no evidence of learning transfer across interactions within the same game (likely due to the lower number of rounds per interaction and the higher cognitive load of facing two opponents rather than one), we found evidence for learning transfer across games as early round scores analysis confirmed. We also found that learning transfer is modulated by the type of opponent faced. When the players faced the level-1 opponent, they were able to transfer learning. However, when they faced the level-2 opponent, there was weaker evidence for transfer. The lack of transfer when facing the more sophisticated opponent might be due to the difficulty of learning that opponent strategy to start with. A player cannot transfer what they have not learnt and as such, since it might be harder to learn the strategy of the level-2 opponent, this in turn might translate into weaker evidence for transfer.

Coming back to evidence for learning transfer, it is undeniable that participants start off new games with prior knowledge as their scores are much higher than chance, confirmed both by early stage analysis as well as rounds 2-6 scores analysis. The question we ask ourselves therefore is: What exactly did the players learn in RPS that allowed them to beat the opponent in FWG and SHOOT? what did the players learn exactly about the opponent strategy and what form did this learning take?

We will proceed by eliminating multiple hypotheses. First, maybe players simply learn spatial heuristics that allow them to perform better than chance. An example is a spatial heuristic that consists of choosing "weapons" in a particular order (for instance left to right). Thsi was one of the weaknesses in the design of experiment one, and we took this into accounr in experiment two by randomly shuffling the spatial order of action choices in each round. Still, the learning and conclusions were similar, so this could not explain both learning and transfer in experiments one and two.

A second possible hypothesis for learning the opponent’s strategy is the use of simple heuristis based on last round play (for instance, I play scissors whenver opponent played rock in last round, or whenever the last round play was rock/scissors, I should play paper in this round, etc...). Our Q-learning with states as prior round play model is a good proxy for this type of strategies. While this approach certainly seemed to be the best fit for player's behavior, it is unsatisfactory in explaining some of the learning transfer evidence we showed. Indeed, learning the best action in a particular state cannot easily transer to a new game since the state space is different and there is no simgle mapping between the state spaces of the initial and latter games. These rules would thereofre need to be learned anew in the latter game which is inconsistent with above chance performance in very early rounds.

Likewise, assuming that players learn a complete model of the environment (for instance the transition probabilities from last round play to new play) might explain learning within games but is equally unable to account for early games transfer of learning as such models, besides being cognitively very expensive to learn, require a long time to learn. Another issue with these hypotheses is that they are not consistent with significant score differences between those facing level-2 and level-1 opponents. After all, if players were using some type of associative learning or spatial heuristics, then their scores should not depend on the degree of strategic sophistication of the opponent since their approaches would render this variable irrelevant. The fact that the degree of sophisitication of the opponent matters points to the importance of opponent modelling to successful transfer of learning.

The more likely explanation of the learning transfer is that it is driven by a group of participants that are able to build a mental representation of what the strategy of the opponent is. A successful mental representation would take the perspective of the opponent or endow it with intentionality in order to detect its strategy when the opponent is playing based on a level-k reasoning model. For instance, the player may think “My opponent is always trying to be one step ahead of me, therefore, I will be one step ahead of where it thinks I will be”.

This mental representation would facilitate the use of theory of mind abilities and thus enable the players to learn opponent strategies when they are based on human-like reasoning models such as level-k or cognitive hierarchy. This type of learning would be deemed “explicit” in the psychology literature as a process through which knowledge consists of cognitive representations of concepts and rules, as well as the relationship between them. It involves the evaluation of explicit hypotheses and results in better problem-solving skills (Mandler, 2004). Since it is less context dependent, this type of learning is generalizable to new situations.

"""

To explain the presence of learning transfer, we believe that the most probable account is that some participants are learning a “strategic rule” by which they can anticipate the opponent’s next action and best respond to it. Certainly, given the heterogeneity in game scores, not all participants were able to do that. As we can see from figure 13, some players are able to score consistently high in early stages and keep a good performance throughout the game while other never quite manage to consistently beat the opponent.

Given the evidence for transfer in early rounds, we believe a significant number of participants did indeed uncover patterns in the opponent’s play that they could leverage to anticipate future actions. In fact, the correlation coefficient between scores against level-1 and level-2 opponents is 57%, significantly higher than 0.

"""
